---
title: "Exploratory Data Analysis"
output:
  pdf_document: default
  html_notebook: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, tidy = TRUE, tidy.opts=list(width.cutoff=60))
```

### Upload the libraries:

```{r}
if (!require("ggplot2", character.only = TRUE)) {
  install.packages("ggplot2")
  library(ggplot2)
}

if (!require("patchwork", character.only = TRUE)) {
  install.packages("patchwork")
  library(patchwork)
}

if (!require("psych", character.only = TRUE)) {
  install.packages("psych")
  library(psych)
}
```

### Upload the data:

```{r}
scales <- read.csv("scales_small_charlson.csv", header= FALSE , stringsAsFactors= FALSE)
lipids <- read.csv("lipids_small_charlson.csv", header= FALSE , stringsAsFactors= FALSE )
colnames(lipids) <- lipids[1,]
lipids <- lipids[-1,]
colnames(scales) <- scales[1,]
scales <- scales[-1,]
scales_lipids <- merge(scales, lipids, by = "MS ID")
```

### Let’s convert the data to the required type:

```{r}
columns_to_convert <- which(names(scales_lipids) != 'sex' & names(scales_lipids) != 'MS ID')
scales_lipids[, columns_to_convert] <- lapply(scales_lipids[, columns_to_convert], as.numeric)
```


# Let's estimate the distribution of demographic factors in the data:

## Age distribution:

```{r}
ggplot(data = scales_lipids, aes(x = age)) +
  geom_histogram(fill = "#B4EEB4", color = "black", bins = 15) +
  labs(x = "Age", y = "Count") +
  theme(axis.title.x = element_text(size = 15),
        axis.title.y = element_text(size = 15))+  
  theme(axis.text.x = element_text(size = 15),
        axis.text.y = element_text(size = 15))
```

### Most patients are between 90 and 95 years old.

## Genders distribution:

```{r}
ggplot(scales_lipids, aes(x = factor(sex))) +
  geom_histogram(aes(fill = factor(sex)), color = "black", bins = 20, stat = "count") +
  scale_fill_manual(values = c("f" = "pink", "m" = "lightblue")) +
  labs(x = "Gender", y = "Count", fill = "Gender") +
  theme(axis.title.x = element_text(size = 15),
        axis.title.y = element_text(size = 15),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15)) +
  theme(axis.text.x = element_text(size = 15),
        axis.text.y = element_text(size = 15)) 
```

### Most patients are female.

## Age distribution in different genders:

```{r}
ggplot(data = scales_lipids, aes(x = age, fill = sex)) +
  geom_density(alpha = 0.5) +
  labs(x = "Age", y = "Density", fill = "Genders") +
  theme(axis.title.x = element_text(size = 15),
        axis.title.y = element_text(size = 15),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15)) +
  theme(axis.text.x = element_text(size = 15),
        axis.text.y = element_text(size = 15)) 
```

### There are differences in the distribution of age between the genders.

# Let's estimate the distribution of scales in the data:

## Distribution of scales by gender:

```{r}
p1 <- ggplot(data = scales_lipids, aes(x = mmse, fill = sex)) +
  geom_density(alpha = 0.5) +
  labs(x = "Mmse", y = "Density", fill = "Sex") +
  ggtitle("Mmse distribution in different genders") +
  theme(plot.title = element_text(hjust = 0.5, size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 12)) +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12)) 

p2 <- ggplot(data = scales_lipids, aes(x = fab, fill = sex)) +
  geom_density(alpha = 0.5) +
  labs(x = "Fab", y = "Density", fill = "Sex") +
  ggtitle("Fab distribution in different genders")+
  theme(plot.title = element_text(hjust = 0.5, size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 12)) +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12)) 

p3 <- ggplot(data = scales_lipids, aes(x = bartel, fill = sex)) +
  geom_density(alpha = 0.5) +
  labs(x = "Bartel", y = "Density", fill = "Sex") +
  ggtitle("Bartel distribution in different genders")+
  theme(plot.title = element_text(hjust = 0.5, size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 12)) +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12)) 

p4 <- ggplot(data = scales_lipids, aes(x = sppb, fill = sex)) +
  geom_density(alpha = 0.5) +
  labs(x = "Sppb", y = "Density", fill = "Sex") +
  ggtitle("Sppb distribution in different genders")+
  theme(plot.title = element_text(hjust = 0.5, size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 12)) +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12)) 

(p1 + p2) / (p3 + p4)
```

### The distribution of the scales differs between genders, especially in the case of the sppb scale.

## Distribution of scales by age:

```{r}
p5 <- ggplot(scales_lipids, aes(x=age, y=mmse)) +
  geom_point(shape=16, 
             size=3, 
             color=adjustcolor("#1874CD"),
             alpha = 0.2) +
  labs(title="Mmse distribution at different ages", x="Age", y="Mmse")+
  theme(plot.title = element_text(hjust = 0.5, size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 12)) +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12)) 

p6 <- ggplot(scales_lipids, aes(x=age)) +
  geom_point(aes(y=fab), 
             shape=16, 
             size=3, 
             color=adjustcolor("Red"),
             alpha = 0.2) +
  labs(title="Fab distribution at different ages", x="Age", y="Fab")+
  theme(plot.title = element_text(hjust = 0.5, size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 12)) +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12)) 

p7 <- ggplot(scales_lipids, aes(x=age)) +
  geom_point(aes(y=bartel), 
             shape=16, 
             size=3, 
             color=adjustcolor("#BF3EFF"),
             alpha = 0.2) +
  labs(title="Bartel distribution at different ages", x="Age", y="Bartel")+
  theme(plot.title = element_text(hjust = 0.5, size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 12)) +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12)) 

p8 <- ggplot(scales_lipids, aes(x=age)) +
  geom_point(aes(y=sppb), 
             shape=16, 
             size=3, 
             color=adjustcolor("#8B1C62"),
             alpha = 0.2) +
  labs(title="Sppb distribution at different ages", x="Age", y="Sppb")+
  theme(plot.title = element_text(hjust = 0.5, size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 12)) +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12)) 

(p5 + p6) / (p7 + p8)
```

# Let's evaluate the correlation between scales and demographic factors:

### Let's convert gender data to a numeric type:

```{r}
names(scales_lipids)[names(scales_lipids) == "sex"] <- "gender"
names(scales_lipids)[names(scales_lipids) == "(-1)Charlson"] <- "charlson"
scales_lipids$gender <- ifelse(scales_lipids$gender == "f", 0, 1)
```

## Let's draw a correlation plot:

```{r}
corPlot(scales_lipids[, c("mmse", "fab", "bartel", "sppb", "age", "gender","charlson" )], 
        cex = 1.2, 
        stars = TRUE,
        alpha = 0.25,
        cex.axis = 0.8,
        main = "Correlations between scales \n and demographic characteristics",
        cex.main = 1)
```

### The scales are highly correlated with each other, but not with demographic factors.
